Autonomous Oceanographic Data Collection towards Estimating Excursion Sets of Vector-valued Random Fields

### 2020-06-12 section2.tex
Updates from Overleaf

\documentclass[aoas]{imsart}

\input{commands.tex}
\usepackage{amsmath}
\usepackage{pstricks,pst-grad}
\usepackage{graphicx}
\usepackage{floatrow}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\floatsetup[table]{capposition=top}
\usepackage{subfigure}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}                     % horizontal lines in tables
\usepackage{comment}

% == Enable text degreehttps://www.overleaf.com/project/5a3268379ecbdc657d8767e8
\usepackage{textcomp}

\usepackage{amsthm,amsmath,natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

% == Trygve Test
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

% put your definitions there:
\startlocaldefs
\newcommand{\edcomment}[1]{{\color{green}{\{Editor: #1\}}}}
\newcommand{\frevcomment}[1]{{\color{blue}{\{Rev 1: #1\}}}}
\newcommand{\srevcomment}[1]{{\color{red}{\{Rev 2: #1\}}}}
\newcommand{\trevcomment}[1]{{\color{violet}{\{Rev 3: #1\}}}}
\endlocaldefs

% New Marcos.
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{cleveref}
\input{macros}

\begin{document}

\begin{frontmatter}

% "New Title of the paper"
\title{Autonomous Oceanographic Data Collection towards Estimating Excursion Sets of Vector-valued Random Fields} 
\runtitle{Autonomous Oceanographic Data Collection}

% "Former Title of the paper" (DG 08.06.2020)
%\title{Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields} \runtitle{Excursion Probabilities for Informative Sampling}

\begin{aug}
\author{\fnms{Trygve Olav} \snm{Fossum}\thanksref{t1,t2}, \corref{} \ead[label=e1]{trygve.o.fossum@ntnu.no}}
\author{\fnms{Cédric} \snm{Travelletti}\thanksref{t3}, \corref{} \ead[label=e2]{cedric.travelletti@stat.unibe.ch}}
\author{\fnms{Jo} \snm{Eidsvik}\thanksref{t4}, \ead[label=e3]{jo.eidsvik@ntnu.no}}
\author{\fnms{David} \snm{Ginsbourger}\thanksref{t3}, \ead[label=e4]{david.ginsbourger@stat.unibe.ch}}
\and
\author{\fnms{Kanna} \snm{Rajan}\thanksref{t5}. \ead[label=e5]{kanna.rajan@fe.up.pt}}

\affiliation[t1]{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU), Trondheim, Norway.} 
\affiliation[t2]{Centre for Autonomous Marine Operations and Systems, NTNU.}
\affiliation[t3]{Institute of Mathematical Statistics and Actuarial Science, University of Bern, Switzerland.}
\affiliation[t4]{Department of Mathematical Sciences, NTNU.}
\affiliation[t5]{Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal.}

\address{\\Trygve Olav Fossum \\Department of Marine Technology\\ Otto Nielsens veg. 10, 7491 Trondheim\\ Norway\\
\printead{e1}}
\address{Cédric Travelletti\\ Institute of Mathematical Statistics and Actuarial Science \\ University of Bern \\
Switzerland.
\printead{e2}}
\address{Jo Eidsvik\\Department of Mathematical Sciences\\ Hogskoleringen 1, 7491 Trondheim\\ Norway\\ \printead{e3}}
\address{David Ginsbourger\\ Institute of Mathematical Statistics and Actuarial Science \\ University of Bern \\
Switzerland.
\printead{e4}}
\address{Kanna Rajan\\Underwater Systems and Technology Laboratory,
  Faculty of Engineering,\\ Rua Dr. Roberto Frias\\ University of Porto, Portugal\\
\printead{e5}}

\runauthor{TO. Fossum et al.}
\end{aug}

\begin{abstract}

  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand processes in the water-column, the
  combination of statistics and autonomous robotics provides new
  opportunities for experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal ocean processes by providing informative descriptions of
  spatial characteristics of ocean phenomena. Specifically, we define
  a design
  criterion based on improved characterization of the uncertainty in
  the excursions of vector-valued Gaussian random fields, and derive
  tractable expressions for the expected Bernoulli variance reduction
  in such a framework. We demonstrate how this criterion can be used
  to prioritize sampling efforts at locations that are ambiguous,
  making exploration more effective. We use simulations to study the
  properties of methods and to compare them with state-of-the-art
  approaches, followed by results from field deployments with an
  autonomous underwater vehicle (AUV) as part of a study mapping the
  boundary of a river plume. The results demonstrate the potential of
  combining statistical methods and robotic platforms to effectively
  inform and execute data-driven environmental sampling.

%Motivated by the challenges related to efficient allocation of sampling resources in environmental sensing, the combination of Excursion Probabilities and Gaussian process modeling is explored for autonomous robotic sampling of ocean features; enabling information driven measures  sampling efforts to high-interest regions. These regions are usually characterized by gradients of measurable environmental variables, e.g., temperature or salinity gradients, on which EPs subsequently can be used...Correlation among samples and multivariate requirements are typical in environmental studies.

\end{abstract}

\begin{keyword}
\kwd{Ocean Sampling}
\kwd{Excursion Sets}
\kwd{Gaussian Processes}
\kwd{Experimental Design}
\kwd{Autonomous robots}
\kwd{Adaptive Information Gathering}
\end{keyword}

\end{frontmatter}
\section{Introduction}

%NB: The subsections are used as a temporary instrument to highlight the proposed structure and should probably be abolished upon convergence to a stable version.

\subsection{Oceanic data collection and spatial design of experiments}

%\textcolor{blue}{NB: I started to match previously listed thoughts/ideas with pieces of the former intro. Will do more asap. Also, Section 3 I find excellent (and more precise than the intro) and start thinking that some bits might end up in the newer intro (while some others could land in a later section).}

Monitoring the world's oceans has gained increased importance in light of the changing climate and increasing anthropogenic impact. 
\begin{comment} % TO BE RE-INSERTED LATER IN THE TEXT?
A central problem for understanding these factors is the lack of representative data with sufficient resolution. Most of this \emph{undersampling} can be attributed to the large spatio-temporal variations in which ocean processes transpire, prompting the need for effective means of sampling. By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use autonomous robotic platforms can be combined with statistical methods to pursue measurements with high scientific relevance. This combination, of statistical tools and robotic platforms, constitute a methodological basis for addressing the problem of efficient sampling of the ocean, which is at the core of this work.
\end{comment}
Central to understanding the changes taking place in the upper
water-column is knowledge of the bio-geophysical interaction driven by
an agglomeration of physical forcings (e.g. wind, topography,
bathymetry, tidal influences, etc.) and incipient micro-biology driven
by planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers and streams.
These often result in a range of ecosystem-related phenomena such as
blooms and plumes, with direct and indirect effects on society. \kc{we
  should look for a citation here. Check with John Ryan}
 One of the bottlenecks in the study of such phenomena lies however in the lack of observational data with sufficient resolution. Most of this \emph{undersampling} can be attributed to the large spatio-temporal variations in which ocean processes transpire, prompting the need for effective means of data collection. 
%
 \textcolor{blue}{By \emph{sampling}, we refer here primarily to the design of observational strategies in the spatial domain %where the use autonomous robotic platforms can be combined with statistical methods 
 with the aim to pursue measurements with high scientific relevance}.
Based on collected data and sometimes also of prior knowledge, models from spatial statistics can be used to make predictions of the state variables under consideration over regions of interest. In case of multiple continuous state variables being jointly studied relying on spatially scattered data, vector-valued random field models and associated prediction approaches such as co-Kriging \citep[See, e.g.,][]{Wackernagel2003} can be considered nowadays as part of the standard statistical toolkit in oceanographic research and beyond. %applications and beyond.  
 %By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use autonomous robotic platforms can be combined with statistical methods to pursue measurements with high scientific relevance. This combination, of statistical tools and robotic platforms, constitute a methodological basis for addressing the problem of efficient sampling of the ocean, which is at the core of this work.

Data collection to understand natural and artificial phenomena at sea has typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}, and also come with computation costs that can be practically limiting. 
%limiting computing resources, no time for hard-core ocean model Navier-Stokes, hence GPs and approximations, etc.
%
\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Traditional ocean observation based on 
    ship-based sampling has been augmented by autonomous
    robotic vehicles. % and their interactions.  
    AUV platforms are an integral part of this network being able to
    reason and make decisions for efficient onboard adaptive sampling.
    % using the sense-plan-act control approach to autonomous control. 
    \ref{fig:nidelven} The interaction of river and ocean creates
    processes that are challenging to map, where the combination of
    statistics and robotics can play a vital role in enabling more
    effective oceanographic observation.\kc{The inset ``AUV robot''
      looks very tacky and ad-hoc; might be best to remove it. Change
      ``ships'' to
      ``Ship'', remove the aerial drone, since we didn't use it and
      don't plan to in this work.}}
  \label{fig:envir} \end{figure}
%
\begin{comment} % Somehow redundant with the parag following below
Mobile robotic platforms \citep{Bellingham07} have contributed significantly to environmental monitoring and sampling. In particular, AUVs have advanced the state of in-situ sampling, and made robotics an integral part of ocean observation, filling parts of the undersampling gap \cite{rudnick03,rudnick18} (See Fig. \ref{fig:envir1}).
\end{comment}
%
The advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean. In particular, autonomous underwater vehicles (AUVs) have advanced the state of data collection and consequently have made robotics an integral part of ocean observation; %our 
previous work by some of the authors of the present paper have contributed to this effort \citep{das11b,Das2015,fossuminformation,fossum18b}.

As full numerical ocean models based on complex differential equations cannot be run onboard a marine robot with limited computational capacity, statistical models relying on collected data and random field assumptions appear quite relevant as means to guide AUV data collection trajectories. Our main research angle in the present work is to extend sequential design strategies from the world of spatial statistics and computer experiments to frameworks featuring both vector-valued observational data and experimental designs that may be constrained by the nature of feasible trajectories.      
\\

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. Missions can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations (sense-plan-act
cycle in Fig. \ref{fig:envir1}), making the information gathering
\emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}.
% This usually occurs on a spatial grid, called a waypoint graph. 
In doing so, the space of sampling opportunities is still limited by
the so-called waypoint graph, a discretization of the search domain, but the AUV has flexibility to modify its path at each waypoint based on in-situ sensing and measurements onboard \citep{py10,Rajan12,Rajan12b}.\\

The work presented here is primarily inspired by a case study 
%application where an AUV is used to measure both temperature and salinity in 
pertaining to the spatial characterization of a frontal system generated by
river plumes. Fig. \ref{fig:nidelven} shows the survey area in
Trondheim, Norway, where cold freshwater enters from the river,
creating a strong gradient in both temperature and salinity. Because
of the local topography and the Coriolis force the cold fresh water
tends to flow to the east. Depending on the river discharge, tidal
effects, wind, and temperature differences, this boundary often gets
distorted. Initial knowledge about the location and evolution of these
features are highly uncertain, making deterministic planning
challenging. Here the goal is to use AUV measurements towards a better description of the interface between freshwater versus oceanic water, with the help of a vector-valued random field model for both temperature and salinity. As we will present in the next sections, some of the original contributions inspired by this specific problem are actually applicable to further settings and application areas. \\

The questions tackled here hence pertain to the broader area of spatial data collection and experimental design, yet with the particularity of estimating some regions of the domain, mostly excursion sets, being implicitly defined by a scarcely observed vector-valued random field. 
%because of limited resources to sample the large oceanographic domains, one must plan for active learning during the operation, where a relevant criterion is used to extract the most valuable designs. The acquired in-situ information must then be assimilated into the statistical model onboard the platform, so that it can be used to re-compute the sampling criterion and in this way inform decisions on where to sample next. \\
Moreover, given constraints on AUV movements and the fact that surveys rely by design in such a framework on successive measurements spread along a trajectory, addressing corresponding design problems calls for particular sequential strategies, possibly also accounting for the finiteness of resources by anticipating multiple steps ahead and thereby improving over myopic strategies. \\

Here we aim to leverage and extent recent progress in uncertainty quantification and reduction for excursion sets of Gaussian random fields in order to address vector-valued observational cases under constrained design such as in the motivating application of designing AUV trajectories to better distinguish between freshwater from ocean waters, relying on temperature and salinity measurements. %approach will take substantial advantage of existing work in the field of excursion set estimation mainly dedicated to computer experiments in the scalar-valued case, and where space exploration can be performed without constraints regarding the distance between successive design points. 
Original contributions compared to recent related work in the framework of scalar-valued (computer) experiments include: 
\begin{itemize}
    \item Extension of two important Sequential Uncertainty Reduction criteria of \citep{Bect.etal2012,chevalier2014fast,bect2019} to homotopic or heterotopic vector-valued settings,
    \item Investigation of myopic and multiple-step ahead batch-sequential strategies for optimizing trajectories with respect to such criteria,
    \item Replicable experiments with synthetic test cases and original data sets with accompagnying codes. 
\end{itemize}
Let us briefly review random field modelling and recent advances in targetd sequential design of experiments based on Gaussian Processes before detailing further existing literature tackling the considered problems, our proposed approach, as well as outlining the rest of the paper.

\subsection{Random field modelling and targeted sequential design of experiments}

 While random field modelling has been one of the main topics throughout the history of spatial statistics \citep{Krige1951a, Matheron1963, Stein1999, Adler.Taylor2007}, there has recently been a renewed interest for random field models in the context of static sequential experimental design, be it in the context of spatial data collection \citep{Mueller2007} or in computer experiments and machine learning \citep[See for instance][and references therein]{Santner.etal2003, Romero.etal2013, Beck2016}. As detailed in \cite{Ginsbourger2018}, Gaussian random field models have been used in particular as a basis to sequential design of comupter experiments dedicated to various goals such as global optimization but also set estimation.  \cite{Picheny.etal2010} revisited the classical Integrated Mean Square Error criterion with an adaptation dedicated to contour line estimation, \cite{Bect.etal2012} focused on strategies to reduce uncertainties on volumes of excursion, i.e. on the measure of the domain of input parameters which associated (scalar) response level exceeds a prescribed threshold, followed up by \cite{chevalier2014fast} presenting computational contributions towards an efficient batch-sequential implementation of strategies introduced in the former. Meanwhile, rather than focusing on excursion volumes, some approaches were investigated in \cite{French.Sain2013,Chevalier.etal2013b,Bolin.Lindgren2015,Azzimonti.etal2016} that ambition to estimate sets themselves. Recently, sequential design of experiments for the conservative estimation of excursion sets based on Gaussian random field models was presented in \citep{Azzimonti.etal}. 
 %
Surprisingly less attention has been dedicated to sequential strategies in the case of vector-valued observations to our knowledge. It has been acknowledge for several decades that co-Kriging models could be efficiently updated with vector-valued observations being assimilated sequentially 
\cite{Vargas-Guzman1999}, sequential strategies targeted to estimate prescribed features of vector-valued random fields are not in still in their infancy. We can refer for instance to \citep{LeGratiet.etal2015} for co-Kriging-based sequential design using fast cross-validation and dedicated to multi-fidelity computer codes. Related ideas are to be found in \cite{Poloczek2017} in the context of Multi-Information Source Optimization. As for extensions of Stepwise Uncertainty Reduction strategies, to our knowledge the only attempt so far to generalize excursion-related Stepwise Uncertainty Reduction criteria (such as reviewed above) in vector-valued settings pertains to the design of multi-fidelity simulations for fire safety, in the PhD thesis \citep[][p.82]{stroh}, where the outputs are mainly assumed independent yet the feasibility of treating the dependent case by appealing to higher-dimensional Gaussian CDF evaluations is mentioned.

% From earlier version; to be relocated?
%Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP) representations of the ocean variables of interest because they are computationally convenient and yet provide enough flexibility to realistically model the spatial variability and dependence, and the correlation between multiple processes. 
\subsection{Contributions, selected previous work on AUV sampling, and outline} 
%work and proposed approach in a nutshell}
%\begin{itemize}
%    \item 
%\end{itemize}
%\textcolor{blue}{We might want to re-use some of the following (from earlier version):}

A major contribution of our work is to derive closed-form results for
this design criteria for situations where the underlying model is
based on multivariate GPs. Another contribution is that we embed this
approach in a sequential strategy to produce sampling algorithms, and
apply this to a real-world scenario of autonomously sampling
temperature and salinity gradients using an AUV, with the intention of
characterizing the variability of oceanic and riverine waters in
typical mixed coastal environments.

Motivating examples relevant for ESs of multivariate processes are
abundant. In medicine, doctors do not rely solely on a single symptom
but must see several combined effects before making a diagnosis. In
our context of environmental sampling, the salinity and temperature
excursions of a river plume can similarly help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}, where sampling can be aimed
towards reducing uncertainty in the temperature and salinity ES.
Unlike what has been done in previous plume exploration studies, we
emphasize the links to spatial statistical modeling and targeted
multivariate sampling criteria for sequential sampling.

Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.
In this paper the focus is not on statistical modeling, but rather on statistical results and computations for efficient sampling designs.

We focus on spatial characterization of a frontal system generated by
river plumes. Fig. \ref{fig:nidelven} shows the survey area in
Trondheim, Norway, where cold freshwater enters from the river,
creating a strong gradient in both temperature and salinity. Because
of the local topography and the Coriolis force the cold fresh water
tends to flow to the east. Depending on the river discharge, tidal
effects, wind, and temperature differences, this boundary often gets
distorted. Initial knowledge about the location and evolution of these
features are highly uncertain, making deterministic planning
challenging.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\cite{bueso1998state,zidek2019monitoring}, but variance and entropy
reduction are independent of the actual data realizations under the
assumptions of GP models, so it has limited
decisional flexibility. The use of data-driven adaptive criteria
was introduced to include more targeted sampling of regions of
scientific interest in \cite{Low2009} and \cite{fossuminformation}. In
this paper, we focus on mapping the river plume by rewarding designs
that improve the classification of ES in temperature and salinity.

\begin{comment}
\subsection{Outline of the paper and its appendices / supplemental material}

\begin{itemize}
    \item 
\end{itemize}
\end{comment}

The remainder of this paper is organized as follows: 
%Section \ref{sec:bg} provides select background on ocean sampling. 
Section \ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for
vector-valued GPs. Section \ref{sec:heuristics} builds on these
assumptions when deriving the sequential design criteria for adaptive
sampling. Section \ref{sec:simulations} discusses properties of the
methods in simulation studies. Section \ref{sec:case_study}
demonstrates the methodology used in field work characterizing a river
plume and finally, Section \ref{sec:concl_disc} contains a summary and
a discussion of future work.

%\subsection{}

%\begin{itemize}
%    \item 
%\end{itemize}

\begin{comment}
\section{Former Introduction (Before 08.06.2020)}

Monitoring the world's oceans has gained increased importance in light
of the changing climate and increasing anthropogenic impact. 
A central problem for understanding these factors is the lack of representative data with sufficient resolution. Most of this \emph{undersampling} can be attributed to the large spatio-temporal variations in which ocean processes transpire, prompting the need for effective means of sampling. By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use autonomous robotic platforms can be combined with statistical methods to pursue measurements with high scientific relevance. This combination, of statistical tools and robotic platforms, constitute a methodological basis for addressing the problem of efficient sampling of the ocean, which is at the core of this work.

The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean. In particular, autonomous underwater vehicles (AUVs), have
advanced the state of sampling and consequently have made robotics an
integral part of ocean observation; our previous work has contributed
to this effort \citep{das11b,Das2015,fossuminformation,fossum18b}.

The potential that lies in using information-based sampling methods, which we
highlight in this paper, is to do efficient environmental sampling and
overcome common challenges as discussed next. First, full numerical ocean models based on complex
differential equations cannot be run onboard a marine robot with
limited computational capacity.
% provide accurate results for online
% operations because the onboard computers lack the computational power
% required.
Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP)
representations of the ocean variables of interest because they are computationally convenient and yet provide enough flexibility to realistically model the spatial variability and dependence, and the correlation between multiple processes. Second, because of limited resources to sample the large oceanographic domains, one must plan for
active learning during the operation, where a relevant criterion is
used to extract the most valuable designs. The acquired in-situ
information must then be assimilated into the statistical model
onboard the platform, so that it can be used to re-compute the
sampling criterion and in this way inform decisions on where to sample
next. We use a criterion focusing on the prediction of excursion sets
(ESs) for the multivariate spatial processes of interest. Third, the
number of sampling designs is enormous, creating a trade-off between
optimization and computability. One must often compromise in both
development and practical operations, and we will rely on
short-horizon optimization routines for sequential designs among
nearest neighbors points in a spatial grid (waypoint graph).

\kc{The para above needs a rewrite to me. It starts off by stating
  there are challenges. But then jumps to models without stating the
  connection with sampling, and then jumps into learning without
  stating why learning is important. Typically the English in formal
  text starts with a statement which is supported or clarified. In the
  two cases above, the connections are simply not obvious. If this was
  there in the first version of the m/s submitted, I'm not sure why
  the reviewers didn't catch it. This is not just a stylistic issue;
  it has to do with making sense and bring the reader/reviewer along.}

\kc{Second, I don't see a motivation for GPs here and as I read
  further into the text. Even for stats folks, are GPs the ``best''
  ways to go? Are they obviously the tools one would use in
  environmental sampling? And even if they are, I think we ought to
  make the case for it strongly and generally, so this paper is
  actually accessible and understandable to those not in the Spatial
  Stats field. I would suggest adding a short para on GPs and why
  they're important -- i.e. motivate the use of GPs early, so we can
  freely make attributions to them later, as in the next section, for
  example.}

The imperfect understanding of causality coupled with the underlying
fundamental uncertainty of the ocean implies that we need to make a
large number of measurements. In addition, the use statistical models
and methods to determine what and how phenomena form, is one way to
increase the quality of predictive power aiding our understanding of
upper water-column processes.

With this in mind, we design AUV sampling strategies for reducing the spatially-integrated Bernoulli variance (IBV) in the ES
indicator variables. 
ES of random fields are thoroughly presented in \citep{adler2009random}. Recent work on ESs and associated excursion
probabilities (EPs) connected to spatial statistics include
\cite{french2013spatio,bolin2015excursion,french2016credible}. Our
focus is on uncertainty characterization and how this is influenced by
sampling. In this sense, our work is similar to
\cite{bect2012,chevalier2014fast,azzimonti2016quantifying} who
describe analytical results for the IBV for univariate processes.

A major contribution of our work is to derive closed-form results for
this design criteria for situations where the underlying model is
based on multivariate GPs. Another contribution is that we embed this
approach in a sequential strategy to produce sampling algorithms, and
apply this to a real-world scenario of autonomously sampling
temperature and salinity gradients using an AUV, with the intention of
characterizing the variability of oceanic and riverine waters in
typical mixed coastal environments.

Motivating examples relevant for ESs of multivariate processes are
abundant. In medicine, doctors do not rely solely on a single symptom
but must see several combined effects before making a diagnosis. In
our context of environmental sampling, the salinity and temperature
excursions of a river plume can similarly help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}, where sampling can be aimed
towards reducing uncertainty in the temperature and salinity ES.
Unlike what has been done in previous plume exploration studies, we
emphasize the links to spatial statistical modeling and targeted
multivariate sampling criteria for sequential sampling.

Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.
In this paper the focus is not on statistical modeling, but rather on statistical results and computations for efficient sampling designs.

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling. Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for
vector-valued GPs. Section \ref{sec:heuristics} builds on these
assumptions when deriving the sequential design criteria for adaptive
sampling. Section \ref{sec:simulations} discusses properties of the
methods in simulation studies. Section \ref{sec:case_study}
demonstrates the methodology used in field work characterizing a river
plume and finally, Section \ref{sec:concl_disc} contains a summary and
a discussion of future work.
\end{comment}

\begin{comment}
\section{Ocean Processes and Observation}
\label{sec:bg}

Central to understanding the changes taking place in the upper
water-column is knowledge of the bio-geophysical interaction driven by
an agglomeration of physical forcings (e.g. wind, topography,
bathymetry, tidal influences, etc.) and incipient micro-biology driven
by planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers and streams.
These often result in a range of ecosystem-related phenomena such as
blooms and plumes, with direct and indirect effects on society. \kc{we
  should look for a citation here. Check with John Ryan}

Traditional data collection to understand such phenomena at sea has
typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}. Mobile robotic platforms
\citep{Bellingham07} have contributed significantly to environmental monitoring and sampling. In particular, AUVs have
advanced the state of in-situ sampling, and made robotics an integral
part of ocean observation, filling parts of the undersampling gap
\cite{rudnick03,rudnick18}(See Fig. \ref{fig:envir1}).

%Walter Munk referred to the 20\textsuperscript{th} century
%as the \emph{century of undersampling} \citep{munk2002}, pointing out
%a lack of sufficient sampling resolution in time and space as a
%critical issue for characterizing ocean phenomena. Even today,
%policymakers struggle with limited observability of severe
%environmental problems; a prominent example is the recent (May 2018)
%outbreak of harmful algal blooms in Northern Norway\footnote{Report
%  from the Institute of Marine Research: https://bit.ly/2KZGJfI}, with
%losses exceeding 12 million tons of salmon valued at \$80 million
%(Norwegian Directorate of Fisheries). This is a case in point of how
%adequate sampling efforts could have provided a more accurate
%assessment and early warning of changing environmental conditions.

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Traditional ocean observation based on 
    ship-based sampling has been augmented by autonomous
    robotic vehicles. % and their interactions.  
    AUV platforms are an integral part of this network being able to
    reason and make decisions for efficient onboard adaptive sampling.
    % using the sense-plan-act control approach to autonomous control. 
    \ref{fig:nidelven} The interaction of river and ocean creates
    processes that are challenging to map, where the combination of
    statistics and robotics can play a vital role in enabling more
    effective oceanographic observation.\kc{The inset ``AUV robot''
      looks very tacky and ad-hoc; might be best to remove it. Change
      ``ships'' to
      ``Ship'', remove the aerial drone, since we didn't use it and
      don't plan to in this work.}}
  \label{fig:envir} \end{figure}

\end{comment}

\begin{comment}
Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. The mission can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations (sense-plan-act
cycle in Fig. \ref{fig:envir1}), making the information gathering
\emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}.
% This usually occurs on a spatial grid, called a waypoint graph. 
In doing so, the space of sampling opportunities is still limited by
the waypoint graph, but the AUV has flexibility to modify its path at
each waypoint based on in-situ sensing and measurements onboard
\citep{py10,Rajan12,Rajan12b}.

We focus on spatial characterization of a frontal system generated by
river plumes. Fig. \ref{fig:nidelven} shows the survey area in
Trondheim, Norway, where cold freshwater enters from the river,
creating a strong gradient in both temperature and salinity. Because
of the local topography and the Coriolis force the cold fresh water
tends to flow to the east. Depending on the river discharge, tidal
effects, wind, and temperature differences, this boundary often gets
distorted. Initial knowledge about the location and evolution of these
features are highly uncertain, making deterministic planning
challenging.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\cite{bueso1998state,zidek2019monitoring}, but variance and entropy
reduction are independent of the actual data realizations under the
assumptions of GP models, so it has limited
decisional flexibility. The use of data-driven adaptive criteria
was introduced to include more targeted sampling of regions of
scientific interest in \cite{Low2009} and \cite{fossuminformation}. In
this paper, we focus on mapping the river plume by rewarding designs
that improve the classification of ES in temperature and salinity.

\end{comment}

\input{section2}

\begin{comment}
\section{Quantifying uncertainty on Excursion Sets implicitly defined by Gaussian processes}
\label{sec:ESEP}

@@

\subsection{Excursion sets and Integrated Bernoulli variance}
\label{sec:ES}

Our objective is to leverage statistical tools onboard an autonomous
robotic platform to characterize a river plume, focusing on spatial
separation of cold freshwater from a river and warmer saline waters
of a fjord. The characterization of uncertainties on the ES is a
useful goal of this sampling strategy.

In order to formalize the problem in mathematical terms, let us
consider a multivariate (i.e., vector-valued) Gaussian random field
$(\gp[\x])_{\x \in \mathcal{M}}$, where $\mathcal{M}$ is an arbitrary
set (typically in geostatistical studies, a geographical subset of interest) and the
dimensionality of the output $\no \geq 1$ is arbitrary. We have $\no = 2$ in the example with temperature and salinity in the river plume.

Assume furthermore that the set of interest
is characterized in terms of a subset of values of the variables, say
$\T \subset \R^{\no}$, and denote the corresponding set of index
points by
$$
\es=\gp^{-1}(\T)=\{\x \in \mathcal{M}: \gp[\x] \in \T\}.
$$
Our goal here is to develop approaches to quantify and improve the
characterization of uncertainties on such sets $\es$, in the
context of multivariate random fields.

>>>>

\section{Quantifying uncertainty on Excursion Sets implicitly defined by Gaussian processes}
\label{sec:ESEP}

@@

Our objective is to leverage statistical tools onboard an autonomous
robotic platform to characterize a river plume, focusing on spatial
separation of cold freshwater from a river and warmer saline waters
of a fjord.

In order to formalize the problem in mathematical terms, let us
consider a multivariate (i.e., vector-valued) Gaussian random field
$(\gp[\x])_{\x \in \mathcal{M}}$, where $\mathcal{M}$ is an arbitrary
set and the dimensionality of the output $\no \geq 1$ is arbitrary.

Assume furthermore that the set of interest
is characterized in terms of a subset of values of the response variables, say
$\T \subset \R^{\no}$, and denote the corresponding set of spatial locations by
$$
\es=\gp^{-1}(\T)=\{\x \in \mathcal{M}: \gp[\x] \in \T\}.
$$
Our goal here is to develop approaches to quantify and improve the
characterization of uncertainties on such sets $\es$, in the
context of multivariate random fields.

While some aspects of the developed approaches do not call for a
specific form of $\T$, we will often stick for simplicity to the case
of orthants
($\T=(-\infty, t_1] \times \dots \times (-\infty, t_{\no}]$ where
$t_1,\dots, t_{\no} \in \R$) as this will allow efficient calculation
of several key quantities. Note that changing some $\leq$ inequalities
to $\geq$ ones would lead to immediate adaptations.
